---
title: "RMS Titanic"
author: "Danny Kennen"
date: 2025-02/27
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class - Where they the 1st, 2nd, 3rd, or crew, class?
-   Sex - Were they Male or Female?
-   Age - Were they an adult or a child?
-   Survived - Did they survive or not?
-   n - the number of people who fall into any given category

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Are there any differences?
    -   The Wikipedia article says there were an estimated 2,224 passengers, compared to the 2,201 that computation said. That's a difference of 23 passengers.
-   If yes, what might account for those differences?
    -   The estimates for passengers and death tolls vary from sources because some passengers traveled under aliases and also records may have

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_survived <- df_titanic %>% filter(Survived == "Yes")
df_survived %>% 
  ggplot() + 
  geom_col(aes(x = Class, y = n, fill = Sex), position = "dodge")
```

**Observations**:

-   The 3rd class was pretty evenly split between females and males who survived.
-   Every class other then crew have more women who survived then men, it's likely that the reason this trend doesn't continue for the crew is because the crew were almost fully men.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_survived <- df_prop %>% filter(Survived == "Yes")
df_survived %>% 
  ggplot() + 
  geom_col(aes(x = Class, y = Prop, fill = Sex), color = "black", position = "dodge")

```

**Observations**:

-   Write your observations here.
    -   The crew's survival rates essentially switch when we measure it proportionally. The majority of the women on the crew survived and the majority of men on the crew died, there were just so few women that even though the majority of the male crew died, the actual number of men who survived is much higher then the women on the crew.
    -   Women in the 3rd class had a higher rate of survival the men (even though over half of them died), there were just likely more men in 3rd class then women so numerically their survival rates look almost even.
-   Is there anything *fishy* going on in your plot?
    -   It says that 100% of both women and men in 1st and 2nd class. This is due to the fact that the graph is overlaying the survival rates for children and adults. When we add the argument color = "black" to our geom_col argument we can see that the columns for adults and children are stacked on top of each other which makes it appear that all women and men survived in first and second class.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_survived <- df_prop %>% filter(Survived == "Yes", Age == "Adult")

df_survived %>%
  ggplot() +
  geom_col(
    mapping = aes(x = Class, y = Prop , fill = Sex),
    position = "dodge")
```

**Observations**:

-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   When we filter out children, we can see that the rates for male survivors goes down on all classes except for crew. The survival rate for females also decreases in every class except for crew, but by less because women were still prioritized much more then men in the evacuation. In this new plot we can see that even women in the 3rd class, who had a less then half chance at survival, still survived at a rate higher then every class of men.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
