---
title: "Antibiotics"
author: "Danny Kennen"
date: 2025-03-09
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------------|------------------------|------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and
supporting files (`report_files/` folder) when you are done! Then submit
a link to Canvas. **Your Challenge submission is not complete without
all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. This means **it must be possible to identify each of the
16 bacteria by name.** You must also show whether or not each bacterium
is Gram positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
df_long_antibio <- df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
               names_to = "Antibiotic",
               values_to = "Concentration_Value"
    ) %>%
  filter(Concentration_Value <= 0.1) %>%
  mutate(bacteria = fct_reorder(bacteria, Concentration_Value)
         )

all_bar_chart <- 
   df_long_antibio %>%
  ggplot(aes(Antibiotic, Concentration_Value, color = gram)) +
  geom_point() +
  facet_wrap(~bacteria) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
all_bar_chart
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. This means **it must be possible to identify each of the
16 bacteria by name.** You must also show whether or not each bacterium
is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
ggplot(df_long_antibio, aes(x = bacteria, y = Concentration_Value, color = gram, shape = Antibiotic)) +
  geom_point() +
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90))
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
df_long_pen <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin),
               names_to = "Pen",
               values_to = "Concentration_Value_pen") %>%
  filter(!is.na(Concentration_Value_pen)) %>%       
  mutate(bacteria = droplevels(factor(bacteria))) 

ggplot(df_long_pen, aes(x = bacteria, y = Concentration_Value_pen)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ gram) +
  scale_y_log10()+
  coord_flip()+
  theme(axis.text.x = element_text(angle = 90))

```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
ggplot(df_antibiotics, aes(penicillin, streptomycin, color = gram, label = bacteria)) +
  geom_point() +
  geom_text_repel(vjust = 1.75, size = 3.5) +
  scale_y_log10() +
  scale_x_log10()
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
df_long_neo_strep <- df_antibiotics %>%
  pivot_longer(cols = c(streptomycin, neomycin),
               names_to = "Antibiotic",
               values_to = "Concentration_Value_neo_strep")

ggplot(df_long_neo_strep, aes(x = Antibiotic, y = bacteria, fill = log10(Concentration_Value_neo_strep))) +
  geom_tile() +
  scale_fill_viridis_c(option = "plasma") 
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

*Observations* - What is your response to the question above?

-   The penicillin has a high negative gram strain and low positive gram
    strain, this means that penicillien is more effective against
    Gram-Postitive bateria. Streptomycin is a relatively even mix of
    negitive and positive gram strains. Neomycin is low in concentration
    for both positive and negative cases, save for a few cases. Both
    Streptomycin and Neoycin both work better on gram-negitive bacteria

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   Visual 3

Why?

-   Visual 3 shows the relationships between the three antibiotics
    between each other, and the comparison of negative and positive gram
    strain cases.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations* - What is your response to the question above?

-   The prefix strepto means chain or twisted, ans diplo means double.
    The Streptococcus pneumoniae virus resembles a chain more then two
    discernible strands. Additionally Diplococcus bacteria reacts
    similarly to the reactions of the *Streptococcus pneumoniae and
    Streptococcus fecalis*

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   Visual 4

Why?

-   The 4th visual shows the concentration of streptomycin and its easy
    to see the concentration for diploccus pneumoniae

Worked with Amir Rosario

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
